Open Data, Open Source and Open Standards in chemistry - The Blue Obelisk five years on

### 2011-04-16 paper.tex
Added text to introduction. Added two references to paper.bib. Added paper.bib, which may have been a mistake.

\author{Charles A Darwin\correspondingauthor$^{1,2}$%
       \email{Charles A Darwin\correspondingauthor - charles@londonzoo.co.uk}%
      \and 
         Egon L Willighagen$^3$%
         \email{Egon L Willighagen - egon.willighagen@ki.se}
     \and 
         Rajarshi Guha$^4$%
         \email{Rajarshi Guha - guhar@mail.nih.gov}
     \and 
         Noel M O'Boyle$^5$%
         \email{Noel M O'Boyle - n.oboyle@ucc.ie}
      \and
         Jane E Doe\correspondingauthor$^2$%
         \email{Jane E Doe\correspondingauthor - jane.e.doe@cambridge.co.uk}
      }

@@

\address{%
    \iid(1)Life Sciences Department, Kings College London, Cornwall House,%
        Waterloo Road, London, UK\\
    \iid(2)Department of Zoology, Cambridge, Waterloo Road, London, UK\\
    \iid(3)Division of Molecular Toxicology, Institute of Environmental Medicine, %
        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden
    \iid(5)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland
}%

@@

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
        \paragraph*{Background:} Text for this section of the abstract.

        \paragraph*{Results:} Text for this section of the abstract \ldots

        \paragraph*{Conclusions:} Text for this section of the abstract \ldots
\end{abstract}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
The Blue Obelisk group was established in 2005 at the
229\textsuperscript{th} National Meeting of the American Chemistry
Society as a response to the lack of open data, open standards and
open source (ODOSOS) in chemistry. While other scientific disciplines
such as physics, biology and astronomy (to name a few) were embracing
new ways of doing science and reaping the benefits of community
efforts, there was little if any innovation in the field of chemistry
and scientific progress was actively hampered by the lack of access to
data and tools.

The formation of the Blue Obelisk group is somewhat unusual in that it
is not a funded network, nor does it follow the industry consortium
model. Rather it is a grassroots organisation, catalysed by an initial
core of interested scientists, but with membership open to all who
share one or more of the goals of the group:
\begin{itemize}
\item Open Source. One can use other people's code without further
permission, including changing it for one's own use and distributing
it again.
\item Open Standards. One can find visible community mechanisms for
protocols and communicating information. The mechanisms for creating
and maintaining these standards cover a wide spectrum of human
organizations, including various degrees of consent.
\item Open Data. One can obtain all data in the public domain when
wanted and reuse it for whatever purpose.
\end{itemize}

>>>>

\author{Charles A Darwin\correspondingauthor$^{1,2}$%
       \email{Charles A Darwin\correspondingauthor - charles@londonzoo.co.uk}%
      \and 
         Egon L Willighagen$^3$%
         \email{Egon L Willighagen - egon.willighagen@ki.se}
     \and 
         Rajarshi Guha$^4$%
         \email{Rajarshi Guha - guhar@mail.nih.gov}
	 \and 
	     Christoph Steinbeck$^6$%
	         \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
     \and 
         Noel M O'Boyle$^5$%
         \email{Noel M O'Boyle - n.oboyle@ucc.ie}
      \and
         Jane E Doe\correspondingauthor$^2$%
         \email{Jane E Doe\correspondingauthor - jane.e.doe@cambridge.co.uk}
      }

@@

\address{%
    \iid(1)Life Sciences Department, Kings College London, Cornwall House,%
        Waterloo Road, London, UK\\
    \iid(2)Department of Zoology, Cambridge, Waterloo Road, London, UK\\
    \iid(3)Division of Molecular Toxicology, Institute of Environmental Medicine, %
        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden\\
    \iid(5)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland\\
    \iid(6)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK
}%

@@

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
        \paragraph*{Background:} The Blue Obelisk movement was established in 2005 as a 
		response to the lack of open data, open standards and
		open source (ODOSOS) in chemistry. While other scientific disciplines
		such as physics, biology and astronomy (to name a few) were embracing
		new ways of doing science and reaping the benefits of community
		efforts, there was little if any innovation in the field of chemistry
		and scientific progress was actively hampered by the lack of access to
		data and tools.

        \paragraph*{Results:} This
		contribution looks back on the past 5 years and surveys progress and
		remaining challenges in the areas of Open Source, Open Data and Open
		Standards in chemistry.

        \paragraph*{Conclusions:} Here we show that the Blue Obelisk has been very successful
		in bringing together researchers and developers with common interests
		in ODOSOS, leading to development of many useful resources freely
		available to the chemistry community. But how best to engage with the
		wider chemistry community outside of the Blue Obelisk remains an open
		question.
\end{abstract}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
The Blue Obelisk movement was established in 2005 at the
229\textsuperscript{th} National Meeting of the American Chemistry
Society as a response to the lack of open data, open standards and
open source (ODOSOS) in chemistry. While other scientific disciplines
such as physics, biology and astronomy (to name a few) were embracing
new ways of doing science and reaping the benefits of community
efforts, there was little if any innovation in the field of chemistry
and scientific progress was actively hampered by the lack of access to
data and tools.

Since 2005 it has become evident that a good amount of development in open 
chemical information and information is driven by the demands of neighbouring
scientific fields. In many areas in biology, for example, the importance of 
small molecules and their interactions and reactions in biological systems
has been realised. In fact, one of the first free and open databases and ontologies 
of small molecules was created as a resource about chemical structure and nomenclature
by biologists \cite{DeMatos:2009p3839}.
Open workflow systems such as Taverna (XXX add citation) have opened the door for 
an easy and seamless integration of FOS components into executable workflows 
or data pipelines. A combination of the Chemistry Development Kit and Taverna, for instance, was
reported in 2010 \cite{Kuhn:2010p4001}.

The formation of the Blue Obelisk group is somewhat unusual in that it
is not a funded network, nor does it follow the industry consortium
model. Rather it is a grassroots organisation, catalysed by an initial
core of interested scientists, but with membership open to all who
share one or more of the goals of the group:
\begin{itemize}
\item Open Source. One can use other people's code without further
permission, including changing it for one's own use and distributing
it again.
\item Open Standards. One can find visible community mechanisms for
protocols and communicating information. The mechanisms for creating
and maintaining these standards cover a wide spectrum of human
organisations, including various degrees of consent.
\item Open Data. One can obtain all data in the public domain when
wanted and reuse it for whatever purpose.
\end{itemize}
